### 2015-09-14 text/ms.Rmd
incorporate comments and edit manuscript

<!--TODO: Jessicas Comments: 
TODO - should the table go in the main text, new fig, grid image?
15. Line 255 from pdf - CMSY predictions were approximately linear  (Fig S4) - Do you mean Fig S4a? Maybe I don't know how to interpret these graphs - I don't see how you can extract that info from the CMSY graph. I don't really know wht you mean by 'high and low prediction clusters'.
16. line 258-59 - if mPRM and SSCOM are high, then true status tends to be high. Is there another example that shows how covariates are exploited. The example shown is convincing in Fig S5, but not in Fig S6l, because the predicted B/Bmsy is only just above (light pink shading). though ths may not matter. you can ignore this comment if you want.
18. line 314-15 need a reference showing that SSCOM contributed little to status from GBM. e.g. which figure shows this?
19. line 334 - should this be "effort dynamics at low biomasses in COM-SIR", not SSCOM? 
20. I said that i'd help with the last paragraph about other uses for ensemble methos. If there is space within the word limit, could we add just one other potential use? Are the references listed in 350-51 using ensemble models or do they simply provide examples of where ensembles could be used. Just check my addition - and delete it if you aren't happy with it. I don't know how to insert references - i'll leave that up to you. 
-->

@@

\begin{Large}
\noindent
Improving estimates of population status and trajectory with superensemble models
\end{Large}

@@

\noindent
Sean C. Anderson^1\*^, 
Andrew B. Cooper^1^,
Olaf P. Jensen^4^, 
C\'{o}il\'{i}n Minto^6^, 
James T. Thorson^11^,
Jessica C. Walsh^1^,
Jamie Afflerbach^2^, 
Mark Dickey-Collas^3,12^, 
Kristin M. Kleisner^5^, 
Catherine Longo^2^, 
Giacomo Chato Osio^7^, 
Daniel Ovando^8^, 
Andrew A. Rosenberg^9^, 
Elizabeth R. Selig^10^

\noindent
\textit{order to be determined and superscripts out of order}

@@

\noindent
^2^National Center for Ecological Analysis and Synthesis, University of
California Santa Barbara, 735 State Street, Suite 300, Santa Barbara, CA 93103,
USA

\noindent
^3^International Council for the Exploration of the Sea, 
H.C. Andersens Boulevard 44-46, DK 1553, Copenhagen, Denmark

\noindent
^4^Institute of Marine & Coastal Sciences, Rutgers University, 
71 Dudley Road, New Brunswick, NJ, 08901-8525, USA

@@

\noindent
^6^Galway-Mayo Institute of Technology, 
Marine and Freshwater Research Centre, Galway, 00000, Ireland

\noindent
^7^EC JRC, IPSC, MAU, Italy

\noindent
^8^University of California Santa Barbara, 
Bren School of Environmental Science and Management, 
Santa Barbara, CA, 93106-5131, USA

\noindent
^9^Union of Concerned Scientists, Cambridge, MA, USA

\noindent
^10^Conservation International, Arlington, VA, USA

\noindent
^11^ Fisheries Resource Analysis and Monitoring Division, Northwest Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, WA 98112

\noindent
^12^DTU Aqua National Institute of Aquatic Resources, Technical University of Denmark (DTU), Jægersborg Alle 1, 2920 Charlottenlund, Denmark

\noindent
^\*^Corresponding author: Sean C. Anderson,
School of Resource and Environmental Management,
Simon Fraser University,
Burnaby BC, V5A 1S6;
E-mail: sean_anderson@sfu.ca

@@

# Abstract

@@

Ecological resource managers are often faced with reconciling multiple potentially conflicting estimates of population status and trajectory. One solution is to average these predictions in an ensemble, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy. Superensemble models, commonly used for climate and weather forecasting, may provide a superior solution. Superensembles use the predictions from multiple models as covariates in a new statistical model. Here we evaluate the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population status and trajectory. We fit four data-limited models of population and exploitation dynamics to marine fish stocks and combine their estimates of the mean and recent trends in biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree. We selected data-limited models for minimal data requirements and thus wide applicability across the majority of the world's fish stocks. We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` stock assessments that incorporated data-intensive stock assessment models. We found that ensemble methods substantially improved estimates of population status and trajectory. Machine-learning superensembles performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`. We found similar improvements when predicting the trajectory of status and when applying the simulation-trained superensembles to catch data for global fish stocks. Ensemble methods have the potential to improve estimates of population status and trends for a wide variety of plant and animal species. However, they must be carefully tested, formed from a diverse set of accurate models, and built on a dataset sufficiently representative of the populations to which they are applied.

\bigskip
\noindent 
Keywords: data-limited fisheries, ensemble methods, multi-model averaging, population dynamics, sustainable resource management
<!--Include on the cover page five to eight words or phrases that will be useful for indexing and literature searches. Do not use words in the title as keywords, and avoid general terms such as conservation.-->

@@

Population status and trajectory are two of the most fundamental values to quantify in conservation biology [e.g. @mace2008; @iucn2015]. These properties are key to everything from global assessments of extinction risk [e.g. @stuart2004; @schipper2008] to assessing the sustainability of hunting [e.g. @fa2002] and fishing exploitation levels [e.g. @worm2009; @juan-jorda2011]. However, conservation and ecological resource managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trajectory [e.g. @brodziak2010; @branch2011; @deroba2015]. Some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.

One solution is to take the average or weighted average of an ensemble of model predictions. Such ensembles are typically more accurate and less biased than individual model estimates and can integrate uncertainty in model structure, initial conditions, and parameter estimation [@dietterich2000; @araujo2007]. This approach forms the basis of many machine learning methods [e.g. @dietterich2000], has helped reconcile climate forecasts from dozens of models [e.g. @murphy2004; @tebaldi2007; @ipcc2013], and even improved early warning signs of malaria outbreaks [@thomson2006]. In ecology, ensemble methods are frequently used to improve species distribution modelling [e.g. @araujo2007; @breiner2015] and indeed have been used to combine estimates of population status and trajectory [e.g. @brodziak2010].

@@

In fisheries science, the commonly used operational model for determining status and trajectory of exploited fish populations are stock assessments, i.e., population models coupled to an observation model, that incorporate all appropriate data (e.g. catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$) [@hilborn1992]. However, the broad range of data required to conduct these stock assessments are not available for the majority of fish populations, including those of conservation concern and of economic interest to fisheries [@fao2014]. Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited data available for the majority of fish stocks: (1) a time series of the total biomass of catch and (2) a basic understanding of population productivity [e.g. @vasconcellos2005; @martell2013]. Recently, @rosenberg2014 investigated the performance of four data-limited models through a large-scale simulation experiment. The four models frequently disagreed about population status and any one model had relatively poor predictive ability on average, but there was evidence suggesting that some models performed better than others depending on circumstances (e.g. Fig. \ref{motivate}).

Here, we estimate population status and trajectory of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models. We apply four ensemble-method approaches of increasing complexity to both simulated and real-world fish stocks and compare their predictive performance against each other and against individual model productions.

@@

<!--
We tested the ability of ensemble methods to improve estimates of population status and trajectory when applied to both a large simulated dataset and a global database of assessed stock status. Here we describe the datasets, individual models of population status, and ensemble methods to combine those estimates. We then describe how we evaluated the ability of the various models to estimate population status.

## Datasets
-->

We first developed and tested ensemble methods on a fully factorial simulated dataset [@rosenberg2014] with known status. These published simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of biomass depletion at the start of the dataset compared to carrying capacity: 100%, 70%, and 40% of carrying capacity; and four exploitation settings: (1) a constant exploitation rate, (2) a exploitation rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where exploitation rate increases continuously, and (4) a 'roller-coaster' scenario where the exploitation rate increases and then decreases. Process noise (recruitment variability; i.e. unexplained variability in population dynamics) was introduced to the models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated or had a first-order autoregressive correlation of $0.6$. The simulation also included a scenario with $N(0, 0.2^2)$ measurement error around log(catch) and one scenario without measurement error.

<!--
TODO: Jamie: add table instead / in addition to this paragraph?
-->

@rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks. The simulation models were built in the package `FLR` [@kell2007] for the statistical software \textsf{R} [@r2015]. The code to generate the simulations is available at <https://github.com/iagomosqueira/stockStatusFAO>.

We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012] --- a compilation of stock-assessment output from hundreds of exploited marine populations around the world. Our analysis of the stock-assessment database was based on version XX downloaded on XX (TODO). After removing stocks for which the models described below did not converge (TODO: add number), this database included `r ram_stocks_n` stocks.
<!-- TODO Dan: Do we address the implications of only running our method on stocks that converged later on? i.e. is there something about those omitted stocks that would make them less suited for this type of analysis?-->

@@

\noindent
where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic population growth rate, $B_0$ represents unfished biomass or carrying capacity $K$, and $C$ represents catch. The fourth model is an empirically derived model based on the RAM Legacy Stock Assessment Database. @rosenberg2014 provide a full summary of these four methods and code to fit all the models is available in an accompanying package `datalimited` (TODO: make public?) for the statistical software \textsf{R}. In summary:

@@

* *mPRM* (modified panel regression model) is a modified version of the panel-regression model from @costello2012. Unlike the other models, mPRM is empirical and not mechanistic --- it uses the RAM Legacy Stock Assessment database to fit a regression model to a series of characteristics of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response. The model used in this paper is modified from the original in that it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 since we do not derive estimates of median status across multiple stocks.

@@

<!--TODO Kristin and Mark: It might be a good idea to add a figure that illustrates the spectrum of ensemble/superensembles that are applied here and what is included (e.g., non-linear relationships, etc.). This will help the reader understand the progression in complexity of the ensemble approaches I think.-->
Superensemble models allow us to incorporate additional covariates and potentially leverage interactions between these covariates and individual model predictions. Additional covariates could be, for example, life-history characteristics, information on exploitation patterns, or statistical properties of the data. For simplicity, and since many possible covariates, such as fishing effort dynamics, would not be known for many real-world data-limited scenarios, we added only one set of additional covariates: spectral properties of the catch time series. Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the statistical properties of the catch series that is independent of time series length (except in affecting precision) and absolute magnitude of catch. We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in \textsf{R} and recorded representative short- and long-term spectral densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year cycles. *TODO add new supp. fig.?*

<!-- TODO KL: I think it's worth explaining that we were interested in how the predictions changed wrt the shape of the real time-series, i.e. how their cyclic patterns of peaks and troughs matched, and that's why other summary statistics, dependent on abundance or length but insensitive to the shape of the trajectory, weren't chosen.-->

@@

The individual models we seek to combine with superensembles provide time series of stock status ($B/B_\mathrm{MSY}$). Therefore, we can use superensembles to estimate multiple properties of these time series. Here, we focus on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years. Together, these quantities address the recent state and trajectory of stock status, which are both of management and conservation interest [e.g. @mace2008]. To avoid undue influence of the end points of the time series on the calculated slope, we measured the slope as the Theil-Sen estimator of median slope [@theil1950].

Here we compare an ensemble average and three superensembles of varying complexity: a linear model with two-way interactions, a random forest, and boosted regression tree. The general form of our models was

$$
\hat{\theta} = 
\argmin_\theta \left( L(b \, | \, \hat{b}_{\mathrm{ensemble}}) \right)
$$
$$
\hat{b}_\mathrm{ensemble} = 
f \left( \hat{b}_\mathrm{CMSY}, \, \hat{b}_\mathrm{COM-SIR}, \,
\hat{b}_\mathrm{SSCOM}, \, \hat{b}_\mathrm{mPRM}, \,
S(0.2), \, S(0.5) \, | \, \theta \right)
$$

\noindent 
where $L$ is a generalized loss function, $f$ is some generalized regression function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY model etc.\ (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $\hat{b}$ are the mean or slope values the model is fitted to, and $\hat{\theta}$ is the set of estimated parameters for the superensemble regression function. The $\argmin_\theta$ refers to finding the set of $\theta$ that minimize the loss function. For all superensemble models of mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the models in log space and exponentiated the predictions. For the estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble models on the natural untransformed scale.
<!--add theta-hat description, b hat, make sure these are right, maybe simplify, describe arg min-->

We fit the linear model superensemble with second-order interactions. We fit the two machine learning superensemble models based on regression trees. Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a specified loss function [@breiman1984]. In random forests, a series of regression trees are built on a random subset of the data and random subset of the covariates of the model [@breiman2001]. In generalized boosted models (GBMs), each subsequent model is fit to the residuals from the previous model; data points that are fit poorly in a given model are given more weight in the next model [@elith2008]. Random forests and GBMs can provide strong predictive performance and fit highly non-linear relationships [@elith2008; @hastie2009]. We fit random forest models with the `randomForest` package [@liaw2002] for \textsf{R} with the default argument values. We fit boosted regression tree models with the `gbm` package [@ridgeway2015] for \textsf{R}. Based on cross-validation with the **caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values.                                                
<!--TODO give more details on tuning or gloss over more-->

@@

A critical component to any predictive modelling exercise is to evaluate the performance of a model on new data [@hastie2009]. We used repeated three-fold cross validation to test predictive performance: we randomly divided the dataset into three sets, built superensemble models on two-thirds of the data, and evaluated predictive performance on the remaining third. We repeated this across each of the three splits and then repeated the whole procedure 50 times to account for bias that may result from any one set of validation splits. In the simulated dataset, there were 10 replicates of each unique combination of simulation parameters that differed only in stochastic variability. Since the dynamics of these populations were often similar, we grouped these stocks in the cross-validation process into either the training or testing split. When applied to the RAM Legacy Stock Assessment database, we used superensembles trained on the entire simulation dataset. However, since mPRM is built on the same stock-assessment database, we refit mPRM on each training split.

@@

<!-- TODO Dan: Should we discuss the issue with MAPE that we ran into before, in that something can have bimodal predictions that are really positively or really negatively biased, and the net result its a MAPE of 0?-->

@@

Applied to the simulated dataset of known stock status, the individual models had variable success at recovering the mean (status) and slope (trajectory) of $B/B_\mathrm{MSY}$ in the last five years. All models exhibited a high degree of scatter around the one-to-one line of perfect status prediction (Fig. \ref{hexagon}). In contrast to the known unimodal distribution of status, CMSY exhibited bimodal predictions (Fig. \ref{hexagon}a), but had the best rank-order correlation and accuracy scores (Fig. \ref{performance}a). COM-SIR and SSCOM both correctly identified a number of stocks with low status, but frequently predicted a high status when status was in fact low (Fig. \ref{hexagon}b, c). mPRM had relatively poor ability to predict status for the simulated dataset (Fig. \ref{hexagon}d). There was generally little correlation between true and predicted trajectory for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) with the exception of SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).

Ensemble methods, and in particular the machine learning superensemble models (random forest and GBM), generally improved estimates of status and trajectory over any individual model (Fig. \ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h). Compared to the individual models, machine learning superensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Fig. \ref{performance}a). These superensembles also had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY} = 1$ (Fig. \ref{roc-sim}). Results were similar when predicting trajectory: compared to individual models, machine learning superensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig. \ref{performance-sim-slope}). The ensemble models that simply took a mean of the individual models ranked slightly behind the best individual model for estimating fish stock status (CMSY; Fig. \ref{performance}a) and had slightly lower correlation but higher accuracy than the best individual model at predicting the trends of status (SSCOM; Fig. \ref{performance-sim-slope}).

>>>>

@@

\begin{Large}
\noindent
Improving estimates of population status and trend with superensemble models
\end{Large}

@@

\noindent
Sean C. Anderson^1\*^, 
Andrew B. Cooper^1^,
Olaf P. Jensen^2^, 
C\'{o}il\'{i}n Minto^3^, 
James T. Thorson^4^,
Jessica C. Walsh^1^,
Jamie Afflerbach^5^, 
Mark Dickey-Collas^6,7^, 
Kristin M. Kleisner^8^, 
Catherine Longo^5^, 
Giacomo Chato Osio^9^, 
Daniel Ovando^10^, 
Iago Mosqueira^9^,
Andrew A. Rosenberg^11^, 
Elizabeth R. Selig^12^

@@

\noindent
^2^Institute of Marine & Coastal Sciences, Rutgers University, 
71 Dudley Road, New Brunswick, NJ, 08901-8525, USA

\noindent
^3^Galway-Mayo Institute of Technology, 
Marine and Freshwater Research Centre, Galway, 00000, Ireland

\noindent
^4^ Fisheries Resource Analysis and Monitoring Division, Northwest Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, WA 98112

@@

\noindent
^6^International Council for the Exploration of the Sea, 
H.C. Andersens Boulevard 44-46, DK 1553, Copenhagen, Denmark

\noindent
^7^DTU Aqua National Institute of Aquatic Resources, Technical University of Denmark (DTU), Jægersborg Alle 1, 2920 Charlottenlund, Denmark

\noindent
^8^National Center for Ecological Analysis and Synthesis, University of
California Santa Barbara, 735 State Street, Suite 300, Santa Barbara, CA 93103,
USA

\noindent
^9^European Commission, Joint Research Centre, Institute for the Protection and
Security of the Citizen, Maritime Affairs Unit G03, Via E. Fermi 2749, 21027
Ispra (VA), Italy

\noindent
^10^University of California Santa Barbara, 
Bren School of Environmental Science and Management, 
Santa Barbara, CA, 93106-5131, USA

\noindent
^11^Union of Concerned Scientists, Cambridge, MA, USA

\noindent
^12^Conservation International, Arlington, VA, USA

\noindent
^\*^Corresponding author, present address: School of Aquatic and Fishery Sciences, Box 355020, University of Washington, Seattle, WA  98195, USA
E-mail: sandrsn@uw.edu

\bigskip
\noindent 
Keywords: data-limited fisheries, ensemble methods, multi-model averaging, population dynamics, sustainable resource management

\bigskip
\noindent 
Word count: TODO

\bigskip
<!--Place the acknowledgment paragraph on the cover page of your manuscript.
(Reviewers are not provided with a cover page.) Do not spell out first (given)
names. Provide the first initial of the first name, even if the initial starts a
sentence. Do not use titles (e.g., Dr. or Professor). Refer to authors of the
manuscript by their initials only (e.g., “S.T.W. was supported by a grant from
the Torry Foundation.”).-->
\noindent
Acknowledgements: We thank members of Phase I of the working group 'INSERT NAME HERE' who
contributed to developing the data-limited methods and simulations used in our
analysis.
We thank J.A. Hutchings for helpful discussions on ensemble methods for
assessing population status.
We thank the Gordon and Betty Moore Foundation for funding of the working group
'INSERT NAME HERE'.
'INSERT OTHER REQUIRED FUNDING ACKNOWLEDGEMENTS HERE'

@@

\begin{LARGE}
\noindent
Improving estimates of population status and trend with superensemble models
\end{LARGE}

@@

\bigskip
\noindent
\textbf{Abstract:} Ecological resource managers must often reconcile multiple conflicting estimates of population status and trend. One solution is to average these predictions in an ensemble, but this approach ignores covariance between models and may not optimally reduce bias and improve accuracy. Superensemble models, commonly used in climate and weather forecasting, may provide a superior solution. Superensembles use the predictions from multiple models as covariates in an additional statistical model fit to known data. Here we evaluate the potential for ensemble and superensemble models ('ensemble methods') to improve estimates of fish population status and trend. We fit four data-limited and widely applicable models of population and exploitation dynamics and combine the models' estimates of the mean and recent trend in biomass relative to the equilibrium biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with an ensemble average and three superensembles: a linear model, a random forest, and a boosted regression tree. We built our models on a simulated dataset of 5760 stocks and tested them with cross-validation and against a global database of `r ram_stocks_n` stock assessments. We found that ensemble methods substantially improved estimates of population status and trend. Machine-learning superensembles (random forest and boosted regression trees) performed the best for estimating population status: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, rank-order correlation between predicted and true status improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`. We found similar improvements when predicting the trend of status and when applying the simulation-trained superensembles to catch data for global fish stocks. Ensemble methods have the potential to improve estimates of population status and trends for a wide variety of plant and animal species. However, they must be tested, formed from a diverse set of accurate models, and built on a dataset representative of the populations to which they are applied.

\bigskip
<!--Include on the cover page five to eight words or phrases that will be useful for indexing and literature searches. Do not use words in the title as keywords, and avoid general terms such as conservation.-->

@@

Population status and trend are two of the most fundamental values to quantify in conservation biology [e.g. @mace2008; @iucn2015]. These properties are key to everything from global assessments of extinction risk [e.g. @stuart2004; @schipper2008] to assessing the sustainability of hunting [e.g. @fa2002] and fishing exploitation levels [e.g. @worm2009; @juan-jorda2011]. However, conservation and ecological resource managers are often faced with reconciling multiple uncertain and potentially conflicting estimates of status and trend [e.g. @brodziak2010; @branch2011; @deroba2015]. Some models may suggest a population is at risk and declining in abundance while others may suggest it is not at risk and stable.

One solution is to take the average or weighted average of an ensemble of model predictions. Such ensembles are typically more accurate and less biased than individual model estimates and can integrate uncertainty in model structure, initial conditions, and parameter estimation [@dietterich2000; @araujo2007]. This approach forms the basis of many machine learning methods [e.g. @dietterich2000], has helped reconcile climate forecasts from dozens of models [e.g. @murphy2004; @tebaldi2007; @ipcc2013], and even improved early warning signs of malaria outbreaks [@thomson2006]. In ecology, ensemble methods are frequently used to improve species distribution modelling [e.g. @araujo2007; @breiner2015] and indeed have been used to combine estimates of population status and trend [e.g. @brodziak2010].

@@

In fisheries science, the commonly used operational model for determining status and trend of exploited fish populations are stock assessments, i.e., population models coupled to an observation model, that incorporate all appropriate data (e.g. catches, size and age distributions, surveys, and tagging information) to quantify values such as the biomass of a stock that can produce maximum sustainable yield ($B/B_\mathrm{MSY}$) [@hilborn1992]. However, the broad range of data required to conduct these stock assessments are not available for the majority of fish populations, including those of conservation concern and of economic interest to fisheries [@fao2014]. Therefore, a number of models have been proposed to assess $B/B_\mathrm{MSY}$ based on the limited data available for the majority of fish stocks: (1) a time series of the total biomass of catch and (2) a basic understanding of population productivity [e.g. @vasconcellos2005; @martell2013]. Recently, @rosenberg2014 investigated the performance of four data-limited models through a large-scale simulation experiment. The four models frequently disagreed about population status, no one model had strong performance across all fish stocks, and some models performed better than others depending on circumstances (e.g. Fig. \ref{motivate}).

Here, we estimate population status and trend of exploited fish populations using ensembles and superensembles (collectively 'ensemble methods') of these four data-limited models. We apply four ensemble-method approaches of increasing complexity to both simulated and real-world fish stocks and compare their predictive performance against each other and against individual model productions.

@@

We first developed and tested ensemble methods on a fully factorial simulated dataset [@rosenberg2014] with known status. These published simulated data included the following scenarios: three fish life histories: small pelagic, demersal, and large pelagic; three levels of initial biomass depletion compared to carrying capacity: biomass at 100%, 70%, and 40% of carrying capacity; and four exploitation settings: (1) a constant exploitation rate, (2) an exploitation rate coupled with biomass to mimic an open-access single-species fishery, (3) a scenario where exploitation rate increased continuously, and (4) a 'roller-coaster' scenario where the exploitation rate increased and then decreased. Process noise (recruitment variability; i.e. unexplained variability in population dynamics) was introduced to the models at two magnitudes in log space, $N(0, 0.2^2)$ and $N(0, 0.6^2)$, and was either uncorrelated or had a first-order autoregressive correlation of $0.6$. The simulation also included a scenario with $N(0, 0.2^2)$ measurement error around log(catch) and one scenario without measurement error.

@rosenberg2014 ran ten iterations for each combination of factors adding stochastic draws of recruitment and catch-recording variability each time to generate a total of 5760 stocks. The simulation models were built in the package `FLR` [@kell2007, <http://flr-project.org>] for the statistical software \textsf{R} [@r2015]. Code to generate the simulations is available at <https://github.com/datalimited/stocksims>.

We also tested our ensemble methods on the RAM Legacy Stock Assessment Database [@ricard2012] --- a compilation of stock-assessment output from hundreds of exploited marine populations around the world. Our analysis of the stock-assessment database was based on version TODO; after removing stocks for which the models described below did not converge (121), this database included `r ram_stocks_n` stocks.

@@

\noindent
where $\hat{B}_{t+1}$ represents predicted biomass at time $t$ plus one year, $B_t$ represents biomass at time $t$, $r$ represents the intrinsic population growth rate, $B_0$ represents unfished biomass or carrying capacity $K$, and $C$ represents catch. The fourth model is an empirically derived model based on the RAM Legacy Stock Assessment Database. @rosenberg2014 provide a full background on these four methods and code to fit the models. In summary:

<!--and code to fit all the models is available in an accompanying package `datalimited` (TODO: make public?) for the statistical software \textsf{R}. In summary:-->

@@

* *mPRM* (modified panel regression model) is a modified version of the panel-regression model from @costello2012. Unlike the other models, mPRM is empirical and not mechanistic --- it uses the RAM Legacy Stock Assessment database to fit a regression model to a series of characteristics of the catch time series and stock with stock-assessed $B/B_\mathrm{MSY}$ as the response. The model used in this paper is modified from the original --- it condenses the life-history categories into three categories to match the simulated dataset, removes the maximum catch predictor since the absolute catch in the simulated dataset is arbitrary, and does not implement the bias correction needed in @costello2012 since we do not derive estimates of median status across multiple stocks.

@@

<!--TODO Kristin and Mark: It might be a good idea to add a figure that illustrates the spectrum of ensemble/superensembles that are applied here and what is included (e.g., non-linear relationships, etc.). This will help the reader understand the progression in complexity of the ensemble approaches I think.-->
Superensemble models allow us to incorporate additional covariates and potentially leverage interactions between these covariates and individual model predictions. Additional covariates could be, for example, life-history characteristics, information on exploitation patterns, or statistical properties of the data. For simplicity, and since many possible covariates, such as fishing effort dynamics, would not be known for many real-world data-limited scenarios, we added only one set of additional covariates: spectral properties of the catch time series. Spectral analysis decomposes a time series into the frequency domain and provides a means of describing the cyclical shape of the catch series that is independent of time series length (except in affecting precision) and absolute magnitude of catch. We fit spectral models to the scaled catch time series (catch divided by maximum catch) with the `spec.ar` function in \textsf{R} and recorded representative short- and long-term spectral densities at frequencies of 0.20 and 0.05, which correspond to 5- and 20-year cycles.

<!--*TODO add new supp. fig.?*-->

@@

The individual models we seek to combine with superensembles provide time series of stock status ($B/B_\mathrm{MSY}$). Therefore, we can use superensembles to estimate multiple properties of these time series. Here, we focus on two properties: the mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years. Together, these quantities address the recent state and trend of stock status, which are both of management and conservation interest [e.g. @mace2008]. To avoid undue influence of the end points of the time series on the calculated slope, we measured the slope as the Theil-Sen estimator of median slope [@theil1950].

For all superensembles, we used the mean or slope of $B/B_\mathrm{MSY}$ as the response and the predictions from the individual models, and the spectral values, as the predictors. For superensemble models of mean $B/B_\mathrm{MSY}$ --- a ratio bounded at zero --- we fit the models in log space and exponentiated the predictions. For the estimates of $B/B_\mathrm{MSY}$ slope, which are not bounded at zero, we fit superensemble models on the natural untransformed scale.

We compared an ensemble average and three superensembles of varying complexity: a linear model with two-way interactions, a random forest, and a boosted regression tree. We fit the linear model superensemble with all second-order interactions. Our two machine learning superensemble models, a random forest and a boosted regression tree, were based on regression trees. Regression trees sequentially determine what value of a predictor best splits the response data into two 'branches' based on a specified loss function [@breiman1984]. In random forests, a series of regression trees are built on a random subset of the data and random subset of the covariates of the model [@breiman2001]. In generalized boosted models (GBMs), each subsequent model is fit to the residuals from the previous model; data points that are fit poorly in a given model are given more weight in the next model [@elith2008]. Random forests and GBMs can provide strong predictive performance and fit highly non-linear relationships [@elith2008; @hastie2009]. We fit random forest models with the `randomForest` package [@liaw2002] for \textsf{R} with the default argument values. We fit boosted regression tree models with the `gbm` package [@ridgeway2015] for \textsf{R}. Based on initial tuning with the **caret** \textsf{R} package [@caret2008], we chose to fit GBMs with 2000 trees, an interaction depth of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other arguments at their default values.

@@

A critical component to predictive modelling is to evaluate the performance of a model on new data [@hastie2009]. We used repeated three-fold cross validation to test predictive performance: we randomly divided the dataset into three sets, built superensemble models on two-thirds of the data, and evaluated predictive performance on the remaining third. We repeated this across each of the three splits and then repeated the whole procedure 50 times to account for bias that may result from any one set of validation splits. In the simulated dataset, there were 10 replicates of each unique combination of simulation parameters that differed only in stochastic variability. Since the dynamics of these populations were often similar, we grouped these stocks in the cross-validation process into either the training or testing split. When applied to the RAM Legacy Stock Assessment database, we used superensembles trained on the entire simulation dataset. However, since mPRM is built on the same stock-assessment database, we refit mPRM on each training split.

@@

@@

Applied to the simulated dataset of known stock status, the individual models had variable success at recovering the mean (status) and slope (trend) of $B/B_\mathrm{MSY}$ in the last five years. All models exhibited a high degree of scatter around the one-to-one line of perfect status prediction (Fig. \ref{hexagon}). In contrast to the known unimodal distribution of status, CMSY exhibited bimodal predictions (Fig. \ref{hexagon}a), but had the best rank-order correlation and accuracy scores (Fig. \ref{performance}a). COM-SIR and SSCOM both correctly identified a number of stocks with low status, but frequently predicted a high status when status was in fact low (Fig. \ref{hexagon}b, c). mPRM had relatively poor ability to predict status for the simulated dataset (Fig. \ref{hexagon}d). There was generally little correlation between true and predicted trend for any of the individual models (rank-order correlation = `r range_slope_corr_non_sscom`) with the exception of SSCOM (correlation = `r slope_corr_sscom`; Figs \ref{scatter-sim-slope}a--d).

Ensemble methods, and in particular the machine learning superensemble models (random forest and GBM), generally improved estimates of status and trend over any individual model (Fig. \ref{hexagon}e--h, Fig. \ref{scatter-sim-slope}e--h). Compared to the individual models, machine learning superensembles improved accuracy (median absolute proportional error) by `r mean_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and reduced bias (median proportional error) from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range` (Fig. \ref{performance}a). These superensembles also generally had better ability to distinguish if simulated stocks were above or below $B/B_\mathrm{MSY} = 0.5$ (Fig. \ref{roc-sim}). Results were similar when predicting trend: compared to individual models, machine learning superensembles improved accuracy by `r slope_sim$mach_vs_ind_mare_fold`%, increased rank-order correlation from `r slope_sim$ind_corr_range` to `r slope_sim$mach_corr_range`, and reduced bias from `r slope_sim$ind_mre_range` to `r slope_sim$mach_mre_range` (Fig. \ref{performance-sim-slope}). The ensemble models that simply took a mean of the individual models ranked slightly behind the best individual model for estimating fish stock status (CMSY; Fig. \ref{performance}a) and had slightly lower correlation but higher accuracy than the best individual model at predicting the trends of status (SSCOM; Fig. \ref{performance-sim-slope}).
